### 2023-02-21 Manuscript/Draft1.Rmd
manuscript update and percent area occ. update

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey. change/add others?**. [[I think these are fine as they are the ones we are concerned with in this paper]]

@@

The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???). To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???). The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios)

@@

General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolut [[absolute]] error and instances where including covarites in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.

@@

In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table. These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.

@@

```{r}

knitr::kable(YT_results, caption = "Yellowtail error results")

knitr::kable(Cod_results, caption = "Cod error results")

knitr::kable(Had_results, caption = "Haddock error results")
```

\section{Discussion}
[[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
number of simulations done for each scenario - time required to do them and analyses
pros and cons of using VAST with covariates in a changing environment
implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
future work - some of the ideas we've discussed already could be mentioned
limitations to our study
perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper
]]

@@

```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library("knitr")
library("kableExtra")
library("dplyr")
library("readr")
library("tidyr")
library("forcats")

result_table = read_csv("CSVs/Yellowtail_results_final.csv")

summary_tab <- result_table %>%

# Group the table according to Parameter, Scenario
  group_by(Scenario,Strata)

# Create table to present the data]
Fig_5 <- kable(summary_tab) %>%
  kable_styling(font_size = 10)  %>%
  pack_rows(tab_kable, colnum = c(1,4),
    index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>%
  pack_rows(tab_kable, colnum = 1,
    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))

```
```{r, results='asis'}
Fig_5
```

@@

param_table1 = read_csv("CSVs/Cod_results.csv")

@@

>>>>

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits

@@

The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios) Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???). To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???). The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios)

@@

General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covarites in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.

@@

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table. These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.

Notes on differences between VAST settings: For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse). 80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings. 24/32 = 75% of scenarios had a VAST run with a better error than the stratified mean. 14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates. Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.

For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings. 28/32~ 88% of scenarios had a VAST run with a better estimate than the stratified mean. All 16/16 = 100% of VAST runs with a covariate were better than the stratified mean. 12/16 VAST were better without covariates. The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise). In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)). Might be having trouble tracking in to smaller strata again.

@@

<!-- # ```{r} -->
<!-- # -->
<!-- # -->
<!-- # knitr::kable(YT_results, caption = "Yellowtail error results") -->
<!-- # -->
<!-- # knitr::kable(Cod_results, caption = "Cod error results") -->
<!-- # -->
<!-- # knitr::kable(Had_results, caption = "Haddock error results") -->
<!-- # ``` -->

\section{Discussion}

Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass). Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions. VAST may provide more consistent biomass predictions(?)

[[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
number of simulations done for each scenario - time required to do them and analyses
pros and cons of using VAST with covariates in a changing environment
implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
future work - some of the ideas we've discussed already could be mentioned
limitations to our study
perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper
]]

@@

<!-- ```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
<!-- library("knitr") -->
<!-- library("kableExtra") -->
<!-- library("dplyr") -->
<!-- library("readr") -->
<!-- library("tidyr") -->
<!-- library("forcats") -->

<!-- result_table = read_csv("CSVs/Yellowtail_results_final.csv") -->

<!-- summary_tab <- result_table %>%  -->

<!-- # Group the table according to Parameter, Scenario -->
<!--   group_by(Scenario,Strata) -->

<!-- # Create table to present the data] -->
<!-- Fig_5 <- kable(summary_tab) %>% -->
<!--   kable_styling(font_size = 10)  %>% -->
<!--   pack_rows(tab_kable, colnum = c(1,4), -->
<!--     index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>% -->
<!--   pack_rows(tab_kable, colnum = 1, -->
<!--     index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))  -->

<!-- ``` -->
<!-- ```{r, results='asis'} -->
<!-- Fig_5 -->
<!-- ``` -->

```{r YTresultsallstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library("knitr")
library("kableExtra")
library("dplyr")
library("readr")
library("tidyr")
library("forcats")

#result_table = read_csv("CSVs/Yellowtail_results_final.csv")
result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper.csv")

summary_tab_all <- result_table %>% 
  subset(Strata == "all")%>%
 # subset(Scenario=="Constant Population")%>%

#   group_by(`Temperature Scenario`)%>%
# # 
# #   group_by(Scenario)%>%
#   mutate(`Temperature Scenario` = factor(`Temperature Scenario`))%>%
#   
#   arrange(`Temperature Scenario`) %>%
  select(Scenario,`Temperature Scenario`,Season,Covariate,Noise,`VAST A`,`VAST B`, `Stratified Mean`)

#COLOR INDIVIDUAL ROWS BY CATEGORY
#define colors

# summary_tab_all <- result_table %>% 
# # Group the table according to Parameter, Scenario
#   group_by(Scenario)%>%
#   mutate(Scenario = factor(Scenario)) %>%
#   arrange(Scenario) %>%
#   select(Scenario, everything())

# Create table to present the data]
Fig_5a <- kable(summary_tab_all[, -1],
      align = c("l", "l", "l", "l", "c", "c", "c"),
  caption = "Yellowtail flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
  row_spec(0,bold = TRUE)%>% #bold column titles
  kable_styling(font_size = 10)  %>%
  pack_rows(tab_kable(), colnum = 1,
    index = table(fct_inorder(summary_tab_all$Scenario), useNA = "no"))%>%
  #color every 4th row a different color
  row_spec(seq(1,96,4), bold = T, color = "white", background = "black")%>%
  row_spec(seq(2,96,4), bold = T, color = "white", background = "#5C5E60")%>%
  row_spec(seq(3,96,4), bold = T, color = "white", background = "#A4A4A4")%>%
  row_spec(seq(4,96,4), bold = T, color = "white", background = "#D2D2D2")

```{r YTresultsreducedstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library("knitr")
library("kableExtra")
library("dplyr")
library("readr")
library("tidyr")
library("forcats")

#result_table = read_csv("CSVs/Yellowtail_results_final.csv")
result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper.csv")

summary_tab_exclude <- result_table %>% 
  subset(Strata == "reduced")%>%
 # subset(Scenario=="Constant Population")%>%

#   group_by(`Temperature Scenario`)%>%
# # 
# #   group_by(Scenario)%>%
#   mutate(`Temperature Scenario` = factor(`Temperature Scenario`))%>%
#   arrange(`Temperature Scenario`) %>%
  select(Scenario,`Temperature Scenario`,Season,Covariate,Noise,`VAST A`,`VAST B`, `Stratified Mean`)

#COLOR INDIVIDUAL ROWS BY CATEGORY
#define colors

Fig_5b <- kable(summary_tab_exclude[, -1],
      align = c("l", "l", "l", "l", "c", "c", "c"),
  caption = "Yellowtail flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
  row_spec(0,bold = TRUE)%>% #bold column titles
  kable_styling(font_size = 10)  %>%
  pack_rows(tab_kable(), colnum = 1,
    index = table(fct_inorder(summary_tab_exclude$Scenario), useNA = "no"))%>%
  #color every 4th row a different color
   row_spec(seq(1,96,4), bold = T, color = "white", background = "black")%>%
  row_spec(seq(2,96,4), bold = T, color = "white", background = "#5C5E60")%>%
  row_spec(seq(3,96,4), bold = T, color = "white", background = "#A4A4A4")%>%
  row_spec(seq(4,96,4), bold = T, color = "white", background = "#D2D2D2");

```{r, results='asis'}
Fig_5a
```
```{r, results='asis'}
Fig_5b
```

@@

#param_table1 = read_csv("CSVs/Cod_results.csv")
param_table1 = read_csv("CSVs/Cod_results_newsettings_forpaper.csv")

@@

```{r HadERROR,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library("knitr")
library("kableExtra")
library("dplyr")
library("readr")
library("tidyr")
library("forcats")

#param_table1 = read_csv("CSVs/Cod_results.csv")
param_table1 = read_csv("CSVs/Haddock_results_newsettings_forpaper.csv")

# Create table to present the data]
Fig_3b <-  kable(summary_tab1[, -1], 
  caption = "Haddock error results.",
    format = "latex", booktabs = TRUE) %>%
  kable_styling(font_size = 10) %>%
  pack_rows(tab_kable(), colnum = 1,
    index = table(fct_inorder(summary_tab1$Temp), useNA = "no"))
```

```{r, results='asis'}
Fig_3b
```
